---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document: 
    keep_md: true
---
## Loading and preprocessing the data

```{r}
setwd("../Activity Monitoring/")
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip","activity.zip")
unzip("activity.zip", "activity.csv")
activity.df = read.csv("activity.csv")
library(data.table)
activity.dt = data.table(activity.df)
activity.dt[,"interval":=as.factor(interval)]
```
## What is mean total number of steps taken per day?

### Calculate the total number of steps taken per day
```{r}
activity.dt[,.(steps.sum=sum(steps,na.rm = TRUE)),by=date]
```

### Histogram of the total number of steps taken each day
```{r}
hist(activity.dt[,.(steps.sum=sum(steps,na.rm = TRUE)),by=date][,steps.sum],main = "histogram of sum of steps per day", xlab = "Total Steps in a Day")
```

### Calculate and report the mean and median of the total number of steps taken per day
```{r}
activity.with.NA.summary = summary(activity.dt[,.(steps.sum=sum(steps)),by=date][,steps.sum])
activity.with.NA.summary
```

## What is the average daily activity pattern?

### Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r}
plot(activity.dt[,.(steps.mean=mean(steps, na.rm = TRUE)), by=interval][,steps.mean],type="l", xlab = "5 minute intervals in a day", ylab = "Average number of steps", main = "Time series plot of average number of steps \n for each 5 minute interval in a day")

interval.max=which.max(activity.dt[,.(steps.mean=mean(steps, na.rm = TRUE)), by=interval][,steps.mean])

steps.max=max(activity.dt[,.(steps.mean=mean(steps, na.rm = TRUE)), by=interval][,steps.mean])

text(interval.max+25,steps.max, labels = paste(activity.dt[interval.max,interval], round(steps.max), sep = ","))
```

## Imputing missing values

### Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}
paste("number of rows with NAs is", sum(is.na(activity.dt[,steps])))
```

### Replace NAs with means  for that 5-minute interval
Strategy for imputing missing values: missing NA values are replaced by the average values for that time interval across the 2 month period

R package "zoo" is used to replace the missing NAs using na.aggregate() function 
```{r}
if(!require("zoo")){
  install.packages("zoo",repos="http://cran.rstudio.com/")
}
library(zoo)
activity.dt.complete = copy(activity.dt)
activity.dt.complete[,steps:=as.integer(na.aggregate(steps)),by=interval]
```

### Histogram of the total number of steps taken each day
```{r}
hist(activity.dt.complete[,.(steps.sum=sum(steps,na.rm = TRUE)),by=date][,steps.sum],main = "histogram of sum of steps per day", xlab = "Total Steps in a Day")
```

### Calculate and report the mean and median of the total number of steps taken per day
```{r}
activity.with.replacement.summary = summary(activity.dt.complete[,.(steps.sum=sum(steps)),by=date][,steps.sum])
activity.with.replacement.summary
```

### Compare means and medians before and after NA replacements
```{r}
boxplot(activity.dt[,.(steps.sum=sum(steps)),by=date][,steps.sum],activity.dt.complete[,.(steps.sum=sum(steps,na.rm = TRUE)),by=date][,steps.sum], names = c("before NA replacement","after NA replacement"), main = "Comparison of Sum of Steps\nbefore and after NA replacement")

points(1:2,c(activity.with.NA.summary["Mean"], activity.with.replacement.summary["Mean"]),pch=15)

text(1:2,c(activity.with.NA.summary["Median"]-1000, activity.with.replacement.summary["Median"]-1500),labels = c(paste("Median = ",activity.with.NA.summary["Median"]),paste("Median = ", activity.with.replacement.summary["Median"])))

text(1:2,c(activity.with.NA.summary["Mean"]+1000, activity.with.replacement.summary["Mean"]+1000),labels = c(paste("Mean = ",activity.with.NA.summary["Mean"]),paste("Mean = ", activity.with.replacement.summary["Mean"])))
```


## Are there differences in activity patterns between weekdays and weekends?

```{r}
activity.dt.complete[,weekday:=ifelse(weekdays(as.Date(date)) %in% c("Saturday","Sunday"),"weekend","weekday")]

activity.dt.complete[,weekday:=as.factor(weekday)]

par(mfrow = c(2,1))

plot(activity.dt.complete[weekday=="weekend",.(steps.mean=mean(steps)),by=interval][,steps.mean], type="l", ylim = c(0,250), xlab = "", ylab = "", main = "Weekend")

plot(activity.dt.complete[weekday=="weekday",.(steps.mean=mean(steps)),by=interval][,steps.mean], type="l", ylim = c(0,250), xlab = "", ylab = "", main = "Weekday")

mtext("------Number of Steps------", side = 2, outer=TRUE, padj = 2)
mtext("------Interval------", side = 1, outer=TRUE, padj = -2)
```